Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated

### 2024-10-23 submission/new_manuscript/manuscript.qmd
revision completed

@@

Advancements in technology have now made it possible to create
near-perfect simulations that are indistinguishable from reality with an
ease, affordability and accessibility that are unprecedented in Human
history.These artificial, yet realistic constructs permeate all areas of
life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

While not all simulations have achieved perfect realism, such as
Computer Generated Images (CGI) in movies or via recent algorithms such
as GANs or diffusion model, which often include distortions or lack
certain key details distinguishing them from real images
[@mcdonnell2010face; @corvi2022detection], it is fair to assume that
these technical limitations will become negligible in the near future.
This is particularly true in the field of face generation, where
face-generation algorithms are already able to create stimuli that are
virtually indistinguishable from real photos [@nightingale2022;
@tucciarelli2020; @moshel2022]. Such a technological feat, however,
leads to a new question: if real and fake stimuli cannot be
differentiated based on their objective "physical" characteristics, how
can we form judgements regarding their nature?

@@

AI-generated content, in particular realistic images, is becoming
commonplace and carries important risks for misinformation and
black-mailing [@viola2023designed], emphasizing the need to understand
the different components that come into play in the formation of reality
beliefs. This exploratory study primarily aims at investigating the
effect of facial attractiveness on simulation monitoring, i.e., on the
beliefs that an image is real or artificially generated. Based on the
*affective reality theory* [@makowski2023affective;
@makowski2018cognitive], which suggests that salient and emotional
stimuli are perceived to be more real (up to a point of reversal after
which beliefs of fiction becomes used an emotion regulation strategy),
we hypothesize a quadratic relationship between perceived realness and
attractiveness: faces rated as highly attractive or unattractive will
more likely be believed to be real. We expect a similar relationship
with trustworthiness ratings given its well-established link with
attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef;
@little2011facial], and a positive relationship with familiarity (as
more familiar faces would appear as more salient, self-relevant and
anchored in reality). Additionally, we will further explore the link
shared by dispositional traits, such as personality and attitude towards
AI, with simulation monitoring tendencies. Importantly, this study does
investigate the discriminative accuracy between "true" photos and "true"
artificially-generated images (which we consider more a technological
issue than a psychological one), focusing on the beliefs that a stimulus
is real or fake, independently of its true. **In other words, the
present study investigates the psychological process that leads to
different *beliefs* of reality, rather than the discrimination between
real faces and actual AI-generated ones, which largely depends on the
technological quality of the AI-generation process.**

@@

All the material (preregistration ***[FOOTNOTE: This approach diverges from the preregistration in several key ways. First, the phrasing of items was modified from "Assuming the face you saw was of a real individual, how..." to "I find this person..." A new attractiveness scale (i.e., Beauty) was introduced to capture a more objective measure of attraction. Finally, the data analysis method was altered from Bayesian Mixed Models due to computational limitations, as we were unable to run these models on a high-performance cluster.]***, experiment demo, experiment code, raw
data, analysis script with complementary figures and analyses, etc.) is
available at
[**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

@@

We excluded 5 participants that either failed 2 (\>= 66.6%) or more
attention check questions, took an implausibly short time to finish the
questionnaires or had incomplete responses. Out of the 5 participants
excluded, 2 Participants were excluded because they failed 2 out of 3
attention checks, 1 because they did not answer the sexual orientation
question, which made further analysis impossible, and 2 had an abnormal
low agreement (r \< 0.1) between the beauty and the attractiveness
ratings (possibly indicating random responses as these two scales
exhibited a higher correlation for the other participants).The final
sample included 145 participants (Mean age = 28.3, SD = 9.0, range:
\[19, 66\]; Sex:48.3% females, 51.0% males, 0.7% others).

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors. \

@@

Several limitations have to be noted. The current experimental paradigm
required participants to judge the realness of faces they had prior
exposure to (which was done to prevent reality judgements from
influencing the other ratings). Although the effect of re-exposure delay
was negligible, the potential bias induced by face familiarity, that is
by re-presenting the same face stimuli twice, as compared to judging
completely new items, cannot be discarded. Future studies could examine
that by incorporating novel face images or increasing the duration of
the re-exposure delay.

Another issue is the impact on reality judgements of the prior explicit
instruction that "about half of the faces were AI-generated and the
other half real photos". Given this prior information given to
participants, it might seems like our enthusiasm pertaining to the
finding that most people did indeed believe a high number of stimuli to
be fake might be unwarranted, since it simply affirms participants
followed the instructions. However, even if that was the case, the
finding that our beliefs of reality can be so easily re-programmed with
simple instructions and lead to high-confidence answers remains an
interesting phenomenon. Moreover, it is to note that the paradigm did
not instruct participants to balance their answers according to a
certain distribution (e.g., 50-50), merely providing them a description
of the dataset. The fact that no presentation order effect was found on
reality beliefs suggests that participants did not try to actively
distribute their responses to match the instructions, in which case we
would have expected a different pattern: for instance, the first few
items judged as real (the initial "true" belief of the participants),
and a bias would progressively appear towards responding "fake" (as
participants realize that all stimuli are of similar nature and that
they have to "make up" for the prevalence of their "real" answers to
fulfil the expected proportion of responses given the instructions).

>>>>

  - name: Ana Neves  
    orcid: 0009-0006-0020-7599

@@

**Advancements in technology have now made it possible to create near-perfect simulations that are indistinguishable from reality with an ease, affordability and accessibility that are unprecedented in Human history**.These artificial, yet realistic constructs permeate all areas of
life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

**While not all simulations have achieved perfect realism, such as Computer Generated Images (CGI) in movies or via recent algorithms such as GANs or diffusion model, which often include distortions or lack certain key details distinguishing them from real images [@mcdonnell2010face; @corvi2022detection], it is fair to assume that these technical limitations will become negligible in the near future.**
This is particularly true in the field of face generation, where
face-generation algorithms are already able to create stimuli that are
virtually indistinguishable from real photos [@nightingale2022;
@tucciarelli2020; @moshel2022]. Such a technological feat, however,
leads to a new question: if real and fake stimuli cannot be
differentiated based on their objective "physical" characteristics, how
can we form judgements regarding their nature?

@@

AI-generated content, in particular realistic images, is becoming
commonplace and carries important risks for misinformation and
black-mailing [@viola2023designed], emphasizing the need to understand
the different components that come into play in the formation of reality
beliefs. This exploratory study primarily aims at investigating the
effect of facial attractiveness on simulation monitoring, i.e., on the
beliefs that an image is real or artificially generated. **The affective reality theory [@makowski2023affective; @makowski2018cognitive] posits that the default tendency is to believe that experiences are real, with emotional and bodily reactions playing a pivotal role in reinforcing or challenging this belief. According to the theory, there is a quadratic (inverse U-shaped) relationship between affect and reality judgments: stimuli that elicit mild to moderate emotional and/or bodily reactions tend to enhance the perception of realness, increasing confidence in reality. However, when emotional or bodily responses become too intense or overwhelming, the default belief shifts towards non-reality beliefs as an emotion regulation mechanism. In other words, extreme emotional reactions can potentially trigger beliefs of reality denial (“it cannot be real”) as a protective mechanism to help individuals cope with distress. The present experiment can be put in relation with the first part of this proposal - that of a positive relationship between embodied or emotional reactions and appraisals of reality. Following this hypothesis, faces rated as either highly attractive or unattractive - and eliciting stronger reactions - would likely be judged as real**. **We expect a similar pattern with trustworthiness, where faces judged as highly trustworthy or untrustworthy will be more likely to be perceived as real. Finally, we anticipate a positive relationship between familiarity and perceived realness, as familiar faces tend to be judged as more real [@miller2023].** Additionally, we will further explore the
link shared by dispositional traits, such as personality and attitude
towards AI, with simulation monitoring tendencies. Importantly, this
study does investigate the discriminative accuracy between "true" photos
and "true" artificially-generated images (which we consider more a
technological issue than a psychological one), focusing on the beliefs
that a stimulus is real or fake, independently of its true. **In other
words, the present study investigates the psychological process that
leads to different *beliefs* of reality, rather than the discrimination
between real faces and actual AI-generated ones, which largely depends
on the technological quality of the AI-generation process.**

@@

All the material (preregistration ***\[FOOTNOTE: This approach diverges
from the preregistration in several key ways. First, the phrasing of
items was modified from "Assuming the face you saw was of a real
individual, how..." to "I find this person..." A new attractiveness
scale (i.e., Beauty) was introduced to capture a more objective measure
of attraction. Finally, the data analysis method was altered from
Bayesian Mixed Models due to computational limitations, as we were
unable to run these models on a high-performance cluster.\]***,
experiment demo, experiment code, raw data, analysis script with
complementary figures and analyses, etc.) is available at
[**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

@@

We excluded 5 participants that either failed 2 (\>= 66.6%) or more
attention check questions, took an implausibly short time to finish the
questionnaires or had incomplete responses. **Out of the 5 participants excluded, 2 participants were excluded because they failed 2 out of 3 attention checks, 1 because they did not answer the sexual orientation question, which made further analysis impossible, and 2 had an abnormal low agreement (r \< 0.1) between the beauty and the attractiveness ratings (possibly indicating random responses as these two scales exhibited a higher correlation for the other participants)**.The final
sample included 145 participants (Mean age = 28.3, SD = 9.0, range:
\[19, 66\]; Sex:48.3% females, 51.0% males, 0.7% others).

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors.\

@@

Several limitations have to be noted. The current experimental paradigm
required participants to judge the realness of faces they had prior
exposure to (which was done to prevent reality judgements from
influencing the other ratings). Although the effect of re-exposure delay
was negligible, the potential bias induced by face familiarity, **that is by re-presenting the same face stimuli twice**, as compared to judging completely new items, cannot be discarded. Future studies could examine
that by incorporating novel face images or increasing the duration of
the re-exposure delay.

Another issue is the impact on reality judgements of the prior explicit
instruction that "about half of the faces were AI-generated and the
other half real photos". Given this prior information given to
participants, it might seems like our enthusiasm pertaining to the
finding that most people did indeed believe a high number of stimuli to
be fake might be unwarranted, since it simply affirms participants
followed the instructions. However, even if that was the case, the
finding that our beliefs of reality can be so easily re-programmed with
simple instructions and lead to high-confidence answers remains an
interesting phenomenon. **Moreover, it is to note that the paradigm did not explicitly instruct participants to balance their answers according to a certain distribution (e.g., 50-50) - merely providing them a description of the dataset (but participants could, and in some cases did, deviate substantially from the information provided).** The fact that no presentation order effect was found on
reality beliefs suggests that participants did not try to actively
distribute their responses to match the instructions, in which case we
would have expected a different pattern: for instance, the first few
items judged as real (the initial "true" belief of the participants),
and a bias would progressively appear towards responding "fake" (as
participants realize that all stimuli are of similar nature and that
they have to "make up" for the prevalence of their "real" answers to
fulfil the expected proportion of responses given the instructions).
